---
title: "04-Enrichment"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages("enrichR")
#install.packages("openxlsx")# independant of java
#install.packages("devtools") 
#devtools::install_github("wjawaid/enrichR")
library(enrichR)
library(openxlsx)

# NOTE - if you have an error "EnrichR website not responding"  and unable to connect to EnrichR database online, try this
# OPTION 1
#devtools::install_github("wjawaid/enrichR")
#library(enrichR)

# OPTION 2 - update curl and wget
# (Step a) Open the "terminal" application for these commands. Follow instructions for mac
# https://help.ubidots.com/en/articles/2165289-learn-how-to-install-run-curl-on-windows-macosx-linux
# 
# (Step b) Install/update wget to the latest version
# Use the same "terminal" application. Enter
# brew install wget
# 
# (step C) Restart your computer. 
# Try out these EnrichR commands to see if you are able to access the enrichR databases online without error message
# install.packages("enrichR")
# library(enrichR)
# enrichR::listEnrichrDbs()

```

## R Markdown - Enrichment anlalysis using EnrichR R package

Summary
- Input gene lists created from T-Test / EdgeR group comparison analysis . Or an short list of genes of interest from your collaborator (Researcher/Clinician/)
- If needed, short list genes based on p-value or FDR value.
- Run EnrichR on short listed gene list

## Import T-test results so that we can short list further
```{r}
#folder that contains group comparison results
fileName <- "input/CRCPilot_Ttest_Shortlisted.csv"

# read in group comparison results
result1 <- read.csv(file = fileName, 
                        header = T, stringsAsFactors = F, row.names = 1)
```

#Filtering/shortlisting example - 1
Short list results based in p-value cut off (pvalue <= 0.05)
```{r}
#pValueCutOff <- 0.05
#which1 <- ((as.numeric(result1$Pvalue) <= pValueCutOff))
#table(which1) #146 genes with p <= 0.01. 719 genes with p-value <= 0.05

#resultShort <- result1[which1, ] #short listed genes 

```

#Filtering/shortlisting example - 2
Short list results based in p-value cut off (pvalue <= 0.01)
```{r}
pValueCutOff <- 0.01
which1 <- ((as.numeric(result1$Pvalue) <= pValueCutOff))
table(which1) #146 genes with p <= 0.01. 

resultShort <- result1[which1, ] #short listed genes 
```

#Filtering/shortlisting example - 3
Short list results based in p-value cut off (p-value <= 0.01), and fold change cut off (greater than 1.5 or less than -1.5), 
```{r , eval=FALSE}
#pValueCutOff <- 0.01
#which1 <- ((as.numeric(result1$Pvalue) <= pValueCutOff) & (as.numeric(result1$SignedFC) >= 1.5 | as.numeric(result1$SignedFC) <= -1.5))
#table(which1) #26 genes TRUE

#resultShort <- result1[which1, ] #short listed genes 
```

#Filtering/shortlisting example - 4
Sort by FDR . 
Take top 500 genes
```{r}
#resultOrder <- result1[order(result1$FDR),] 
#resultShort <- resultOrder[1:500, ]
```

#Filtering/shortlisting example - 5
Short list results based in adjusted p-value cut off (also called q-value or fdr) <= 0.05
```{r}
#qValueCutOff <- 0.05
#which1 <- ((as.numeric(result1$FDR) <= qValueCutOff))
#table(which1) #
#resultShort <- result1[which1, ] #short listed genes
```


## Export the short listed results for reference
Make sure folder "output" exists
```{r}
#length of short listed results
nrow(resultShort)

write.csv(x = resultShort, file = "output/TTest_results_shortlist2.csv")
```

## Clean gene names
Note - the gene names are in the form of "reporterid_genename". So need to split this.
```{r}
funcSplit <- function(rep_gene) {
    rep_gene_split <- unlist(strsplit(x = rep_gene, 
                                      split = "_", 
                                      fixed = TRUE))
    gene <- rep_gene_split[2]
    return(gene)
}
geneListSplit <- apply(X = as.matrix(row.names(resultShort)), 
                       MARGIN = 1, FUN = funcSplit )
head(geneListSplit) #cleaned gene names

#remove duplicates
geneListSplit1 <- unique(geneListSplit)

# remove NA value
geneListFinal <- na.omit(geneListSplit1) #701 genes

#print number of unique genes
length(geneListFinal)
write.table(x = geneListFinal, 
          file = "output/shortListedUniqueGenes.tsv",
          quote = F, sep = "\t", row.names = FALSE, col.names = F)


```

## Load Databases for Enrichr R package , and check connection
```{r}
#checking if EnrichR website and packing are working
#dbs <- enrichR::listEnrichrDbs()  #total number of databases available = 200+ 

#testing if EnrichR package is working
testOutput <- enrichR::enrichr(genes = c("Runx1", "Gfi1", "Gfi1b", "Spi1", "Gata1", "Kdr"), databases = "KEGG_2021_Human")
head(testOutput[[1]])

#List of databases for which enrichment analysis will be run
dblist1 <- read.csv(file = "input/2023-EnrichR-Databases.txt", 
                    header = F, stringsAsFactors = F)

head(dblist1)
```

## call function to run Enrichment

```{r}
# set output file name
outputFileName1 <- paste("output/ColonCancer", "_EnrichR.xlsx", sep="")

#Load R script into the environment
source(file = "functionEnrichment.R") 

#call function to run Enrichment
functionEnrichment(dblist1, geneListFinal, outputFileName1)

#NEED INTERNET CONNECTION
```

Note - you will need internet connection to complete the above step. 

### RESULT
When you run the funtion, will create an Excel file with multiple tabs. Each tab shows the enrichment results for a database. A total of 14-15 databases, so 14-15 tabs in the excel sheet

After you finish coding, click on "Knit" to create an HTML or PDF file with text and code
