---
title: 'Homework 7: Enrichment Analysis in R'
author: "Aizhan Uteubayeva (netID:au198)"
date: "2023-10-11"
output:
  pdf_document: default
  html_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(enrichR)
library(openxlsx)
```

## R Markdown - Enrichment analysis using EnrichR R package

Summary - Input gene lists created from T-Test / EdgeR group comparison analysis . Or an short list of genes of interest from your collaborator (Researcher/Clinician/) - If needed, short list genes based on p-value or FDR value. - Run EnrichR on short listed gene list

## Import T-test results for further shortlisting

```{r}
#folder that contains group comparison results
fileName <- "input/TTest_HW_EGradeIII-IV_(Comp).vs._EGradeI-II_(Base).TTest.csv"

# read in group comparison results
result1 <- read.csv(file = fileName, 
                        header = T, stringsAsFactors = F, row.names = 1)
```

## Filtering/shortlisting

Short list results based in p-value cut off (p-value \<= 0.01), and fold change cut off (greater than 1.5 or less than -1.5) -à refer to R code done in class

```{r}
pValueCutOff <- 0.01
which1 <- ((as.numeric(result1$Pvalue) <= pValueCutOff) & (as.numeric(result1$SignedFC) >= 1.5 | as.numeric(result1$SignedFC) <= -1.5))
table(which1) #82 genes TRUE

resultShort <- result1[which1, ] #short listed genes 
```

## Export the short listed results for reference

```{r}
# length of short listed results
nrow(resultShort)

write.csv(x = resultShort, file = "output/HCC_TTest_results_shortlist2.csv")
```

## Clean gene names

Genes are in the name contain "\|"

```{r}
funcSplit <- function(rep_gene) {
    rep_gene_split <- unlist(strsplit(x = rep_gene, 
                                      split = "|", 
                                      fixed = TRUE))
    gene <- rep_gene_split[2]
    return(gene)
}
geneListSplit <- apply(X = as.matrix(row.names(resultShort)), 
                       MARGIN = 1, FUN = funcSplit )
head(geneListSplit) #cleaned gene names

#remove duplicates
geneListSplit1 <- unique(geneListSplit)

# remove NA value
geneListFinal <- na.omit(geneListSplit1)

#print number of unique genes
length(geneListFinal)
write.table(x = geneListFinal, 
          file = "output/shortListedUniqueGenes.tsv",
          quote = F, sep = "\t", row.names = FALSE, col.names = F)


```

## Loading Databases for EnrichrR package

```{r}
#checking if EnrichR website and packing are working
#dbs <- enrichR::listEnrichrDbs()  #total number of databases available = 200+ 

#testing if EnrichR package is working
testOutput <- enrichR::enrichr(genes = c("Runx1", "Gfi1", "Gfi1b", "Spi1", "Gata1", "Kdr"), databases = "KEGG_2021_Human")
head(testOutput[[1]])

#List of databases for which enrichment analysis will be run
dblist1 <- read.csv(file = "input/2023-EnrichR-Databases.txt", 
                    header = F, stringsAsFactors = F)

head(dblist1)
```

## Calling function to run Enrichment

```{r}
# set output file name
outputFileName1 <- paste("output/Aizhan_Uteubayeva", "_EnrichR_HW.xlsx", sep="")

#Load R script into the environment
source(file = "input/functionEnrichment.R") 

#call function to run Enrichment
functionEnrichment(dblist1, geneListFinal, outputFileName1)
```
